Pool Model - A unified Mathematical\\ Modelling Framework for Food Science

### 2025-09-04 paper/pool_model.tex
remvoe marker
remove todo
adapt formulations of first section

Consider for simplicity the system where only one bacteria species is present $N=1$.
The bacterial growth curves usually consist of different consecutive phases~\cite{buchanan_when_1997}.
At the beginning no increase in bacterial count is present and so-called lag-phase is observed.
This phenomenon appears due to adaptation of the microbial community, for example, to the new environmental conditions.
This so-called preparation to the exponential growth includes synthesis of the needed compounds, initiation of the transcription~\cite{rolfe_lag_2012} and poorly studied physiological and regulatory mechanisms~\cite{monod_growth_1949}.
After this, the succeeded in adaptation bacteria enter the exponential growth phase, where the process of cell division is directly happening.
When the nutrition is exhausted by the growing bacteria, the stationary state of the growth curve occurs.
In this state the cultures accumulate waste products of metabolic activity that leads to cell death of the bacteria.

@@

As a result the balance between the growing and dying processes leads to the constant with time bacterial count~\cite{navarro_llorens_stationary_2010}.
In the described simplified version of the real-life system three possible bacteria states are mentioned corresponding to which process is happening: adaptation, growth or death.
Motivated by this we divide a bacterial population into three pools (states)~(\ref{fig:SchematicRep}) defining the fraction of the population in the lag phase $L(t)$, growing and dividing $G(t)$ and undergoing cell death $D(t)$.

@@

To determine the system of equations describing the system dynamics consider the flows between different states.

@@

For now, we assume for simplicity that there is no back flow between growth state $G$ and lag-state $L$.

@@

\end{split}\end{align}.

@@

According to the works of Jacques Monod, in the system where all needed nutrients are in excess except for one the total growth is linear to the concentration of this limiting nutrient~\cite{monod_growth_1949}.
The control of the population density with respect to the available nutrition is implemented using cell communication process called quorum sensing.
Using release of the extracellular signalling molecules (autoinducer molecules) it is possible to synchronize the gene expression of the whole group~\cite{ng_bacterial_2009}.
We would not dive into the processes of quorum sensing and try to keep it as simple as possible.
According to this, we aim to include a resource pool which is used by the growing bacteria using the second order kinetic reaction:

@@

For simplicity, we rather make a transition to the dimensionless constant $\alpha:=\alpha_0 N_t$ and dimensionless resource pool $R(t) = R_0 / N_t$ ($0 \leqslant R \leqslant 1$).

@@

Such model can be simplified by exploiting that limitation of the resource pool also limits the total bacterial count of the system:

@@

        The blue curve shows significant dip in behavior of the system where the death of the bacteria is present $\mu = 0.5$.
        Other curves show the case of no bacteria dying $\mu=0$ and represent different length of lag phases:
        short $\lambda=10^{-2}$ (dark green), intermediate $\lambda=10^{-3}$ (light green) and long one $\lambda=10^{-4}$ (yellow).

@@

Here we can see that the value of the transition constant $\lambda$ determines how long the bacteria is in lag-phase.
The larger is the rate the shorter is the lag-time.
Moreover, by including the bacteria flow to the death pool we can diversify the shape of the growth curve allowing the decrease of the bacterial count.

@@

Note that one could further simplify this model to a two pool or state model by omitting the pool $D$ and removing cells with rate $\beta$ directly from pool $L$.

@@

The dynamics of the three pool model can be captures by the coupled system of ordinary differential equations (the two pool model can be achieved by setting $D\equiv 0$ and replacing $\mu$ with $\beta$):
to be solved with the initial conditions $L(0)=N_0$, $G(0)=0$, and $D(0)=0$, where $N_0$ is the initial bacterial population.

@@

For constant parameters $\mu$, $\lambda$, and $\beta$ the integrals for $L$ and $D$ can be put into the differential equation for $G$.
However, for time dependent parameters, e.g., for a dependence on a non-static, dynamic temperature, it appears to be not advantageous to use the analytical results for $L$ and $D$ and work instead with all three equations for $L$, $G$, and $D$.

@@

with the initial conditions $G(0)=0$ and $N(0)=N_0$. 
These equations have a clear interpretation, are derived from simple principles, and are different from the popular Barayni-Roberts and modified-Gompertz models.

@@

The equation exhibit a lag phase and an initial drop due to cell death.

@@

Assuming $L\ll N_t$ we can ignore the non-linear term in the equation for $G$, which gives rise to:

@@

For $\alpha > \mu+\lambda$ $t_L$ can be approximated by:

@@

Due to its life in rather unfavorable conditions bacteria developed different surviving mechanisms, one of which is adapting their metabolism and transitioning into dormant state.
By definition dormancy in bacteria is a reversible state where cell continue living with low metabolic activity while the cell division stops~\cite{kaprelyants_dormancy_1993}.

@@

The models including transition of bacteria to the dormant state were proposed and studied by Bär \etal~\cite{bar_modelling_2002} and Jones and Lennon~\cite{jones_dormancy_2010}.

@@

We explore now the idea that under certain conditions, as sudden change in the environment, part of the population $G$ will enter a lag-phase, i.e., there is a back-flow from $G$ to $L$.

@@

One of the reasons of the dormant state appearance can be also the temperature shift~\cite{oliver_viable_1995}.
Moreover, the rapid shift in temperature is associated with the appearance of an additional lag phase~\cite{zwietering_modeling_1994}.

@@

As it was mentioned above, bacteria usually coexist in large multispecies interacting communities whose growth is vastly determined by the developed survival mechanisms.
These survival mechanisms mostly aim to secure as much available nutrition which lead to competitive strategies or, if it is advantageous, cooperative ones~\cite{hibbing_bacterial_2010, stubbendieck_bacterial_2016}.

@@

    % TODO make this into 2x2 grid for plots and insert cases for psi from before as well
    %\includegraphics[width=\columnwidth]{Figures/pool_model_1pool_inhib.pdf}

@@

\textbf{!Christian extend this!}

>>>>

We begin our discussion by considering a simplistic system where only one bacteria species is present $N=1$.
Bacterial growth undergoes different consecutive phases~\cite{buchanan_when_1997}, starting with the so called lag-phase, during which no increase in the bacterial count occurs.
This phenomenon appears due to adaptations of the microbial community, for example, to the new environmental conditions.
This so-called preparation before the exponential growth includes synthesis of the needed compounds, initiation of the transcription~\cite{rolfe_lag_2012} and poorly studied physiological and regulatory mechanisms~\cite{monod_growth_1949}.
After this, the bacteria which have succeeded in this adaptation stage enter the exponential growth phase, where the process of cell division is predominant.
When the levels of nutrition are exhausted by the growing bacteria, the stationary state of the growth curve occurs.
In this state the cultures accumulate waste products of metabolic activity that lead to cell death in the bacterial colony.

@@

As a result the balance between the growing and dying processes leads to a constant bacterial count~\cite{navarro_llorens_stationary_2010}.
In the described simplified version of the real-life system three possible bacterial states are mentioned, corresponding to the underlying process: adaptation, growth or death.
Motivated by this we divide a bacterial population into three pools (states)~(\ref{fig:SchematicRep}), defining the fraction of the population in the lag phase $L(t)$, growing and dividing $G(t)$ and undergoing cell death $D(t)$.

@@

To determine the system of equations which describe the systems dynamics, we consider the flows between the different states.

@@

For now, we assume for that there is no back flow between growth state $G$ and lag-state $L$.

@@

with the initial conditions $L(0)=N_0$, $G(0)=0$, and $D(0)=0$, where $N_0$ is the initial bacterial population.

@@

According to the works of Jacques Monod, in the system where all needed nutrients are in excess except for one, the total growth is linear to the concentration of this limiting nutrient resource~\cite{monod_growth_1949}.
The control of the population density with respect to the available nutrition is implemented using cell communication processes called quorum sensing.
By releasing extracellular signalling molecules (autoinducer molecules) it is possible for the bacteria to synchronize the gene expression of the whole group~\cite{ng_bacterial_2009}.
We will not dive into the processes of quorum sensing and try to keep it as simple as possible.
We aim to include a resource pool which is consumed by the growing bacteria using the second order kinetic reaction:

@@

For simplicity, we make a transition to the dimensionless constant $\alpha:=\alpha_0 N_t$ and dimensionless resource pool $R(t) = R_0 / N_t$ ($0 \leqslant R \leqslant 1$).

@@

Such a model can be simplified by exploiting that any limitation of the resource pool also leads to a limit of the total bacterial count

@@

        The blue curve shows a significant dip in the behavior of the system where the death of the bacteria is present $\mu = 0.5$.
        Other curves show the case of no bacteria dying $\mu=0$ and represent different lengths of lag phases: short $\lambda=10^{-2}$ (dark green), intermediate $\lambda=10^{-3}$ (light green) and long one $\lambda=10^{-4}$ (yellow).

@@

Here we can see that the value of the transition constant $\lambda$ determines how long the bacteria stays in its lag-phase.
The larger this rate, the shorter is the lag-time.
Moreover, by including the flow of living bacteria to the death pool, we can model a greater variety of effects by allowing a decrease in the bacterial count.

@@

Note that the previously introduced model could be further simplified to a two pool or state model by omitting the pool $D$ and removing cells with rate $\beta$ directly from pool $L$.

@@

% The dynamics of the three pool model can be captured by the coupled system of ordinary differential
The dynamics of the three pool model is captured by equations~\ref{eq:ode_3pool_resource2} which can be used to derive the two pool model by setting $D\equiv 0$ and replacing $\mu$ with $\beta$).

@@

For constant parameters $\mu$, $\lambda$, and $\beta$ the integrals for $L$ and $D$ can be inserted into the differential equation for $G$.
However, in the case of time dependent parameters, e.g., for a dependence on a non-static, dynamic temperature, it appears to be not advantageous to use the analytical results for $L$ and $D$ and work instead with all three equations for $L$, $G$, and $D$.

@@

with initial conditions $G(0)=0$ and $N(0)=N_0$.
These equations have a clear interpretation, are derived from first principles, and are not identical to the popular Barayni-Roberts and modified-Gompertz models.

@@

The equation exhibits a lag phase and an initial drop due to cell death.

@@

Assuming $L\ll N_t$, we can ignore the non-linear term in the equation for $G$, which gives rise to:

@@

For $\alpha > \mu+\lambda$  we can approximate $t_L$ by:

@@

Bacteria which have lived under rather unfavorable conditions have developed different surviving mechanisms, one of which is adapting their metabolism and transitioning into dormant state.
Dormancy in bacteria is a reversible state where cell continue living with low metabolic activity while the cell division stops~\cite{kaprelyants_dormancy_1993}.

@@

Multiple models which include transitions of bacteria to the dormant state were proposed by Bär \etal~\cite{bar_modelling_2002} and Jones with Lennon~\cite{jones_dormancy_2010}.

@@

We explore now the idea that under certain conditions, a sudden change in the environment can lead to parts of the population $G$ entering a lag-phase.
Within the framework of our pool model this means, that there is a back-flow from $G$ to $L$.

@@

One of the reasons that bacteria enter the dormant state can be due to a shift in temperature~\cite{oliver_viable_1995}.
Moreover, a rapid shift in temperature is associated with the appearance of an additional lag phase~\cite{zwietering_modeling_1994}.

@@

As mentioned above, bacteria usually coexist in large interacting communities of multiple species whose growth is vastly determined by the developed survival mechanisms.
These mechanisms mostly aim to secure as much available nutrition which lead to competitive strategies or, if it is advantageous, cooperative ones~\cite{hibbing_bacterial_2010, stubbendieck_bacterial_2016}.

@@

@@
